---
title: "HemRemSapRecruit"
author: "Audrey Barker Plotkin"
date: "2026-02-05"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
options(scipen=999)
knitr::opts_knit$set(root.dir = 'C:/Users/aab133/Harvard University Dropbox/Audrey Barker Plotkin/MaterialLegaciesKopecky')
```

TO DO:  if applicable, integrate residualisation code from here: https://drive.google.com/drive/u/0/folders/1olAJZAbpfTRCUqtd82mS0zjUb1haOHxc

Data sets: trees; saplings; dead wood. each summarized by plot, year, and group (e.g. species; deadwood type). Possibly also include litterfall (annual), air temperature (summarized by season, year, ave, other??), light (leaf-on)

Step 1a. Load the tree data & clean it up
Goal: pull out the recruits

Download tree data from EDI, using the EDI's code generation in R tab. This code is copied from there
I subset this to read in only 126-02, the tree data file (there are 4 datasets total, but the other 3 aren't needed here)
```{r}
# Package ID: knb-lter-hfr.126.19 Cataloging System:https://pasta.edirepository.org.
# Data set title: Overstory Vegetation in Hemlock Removal Experiment at Harvard Forest since 2003.
# Data set creator:  Aaron Ellison -  
# Data set creator:  Audrey Barker Plotkin -  
# Contact:    - Information Manager Harvard Forest  - hf-im@lists.fas.harvard.edu
# Stylesheet v2.16 for metadata conversion into program: John H. Porter, Univ. Virginia, jporter@virginia.edu      
# Uncomment the following lines to have R clear previous work, or set a working directory
# rm(list=ls())      

# setwd("C:/users/my_name/my_dir")       


options(HTTPUserAgent="EDI_CodeGen")
	      
inUrl2  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-hfr/126/19/1d1ce143fc74f7f5d96bcd413425d149" 
infile2 <- tempfile()
try(download.file(inUrl2,infile2,method="curl",extra=paste0(' -A "',getOption("HTTPUserAgent"),'"')))
if (is.na(file.size(infile2))) download.file(inUrl2,infile2,method="auto")

                   
 dt2 <-read.csv(infile2,header=F 
          ,skip=1
            ,sep=","  
        , col.names=c(
                    "tree",     #metadata mismatch has first 3 columns as plot, sec, tree whereas data cols are tree, plot, sec. oops. will fix on my end but this is a workaround
                    "plot",     
                    "sec",     
                    "species",     
                    "dbht0",     
                    "cc",     
                    "condt0",     
                    "yeart0",     
                    "xcoord",     
                    "ycoord",     
                    "zelev",     
                    "elev0",     
                    "markmanip",     
                    "actualmanip",     
                    "cond06",     
                    "dam06",     
                    "sprt06",     
                    "notes1",     
                    "notes06",     
                    "dbh09",     
                    "cond09",     
                    "dam09",     
                    "sprt09",     
                    "notes09",     
                    "hwab109",     
                    "hwab209",     
                    "dbh14",     
                    "cond14",     
                    "dam14",     
                    "sprt14",     
                    "vigor14",     
                    "hwab114",     
                    "hwab214",     
                    "scale14",     
                    "note14",     
                    "dbh19",     
                    "cond19",     
                    "dam19",     
                    "sprt19",     
                    "vigor19",     
                    "hwab119",     
                    "hwab219",     
                    "scale19",     
                    "note19",     
                    "dbh24",     
                    "cond24",     
                    "dam24",     
                    "sprt24",     
                    "vigor24",     
                    "hwab124",     
                    "hwab224",     
                    "scale24",     
                    "note24"    ), check.names=TRUE)
               
unlink(infile2)

# ABP: I'm skipping this part because I do not want so many of the variables to be factors		    
# Fix any interval or ratio columns mistakenly read in as nominal and nominal columns read as numeric or dates read as strings
                
# if (class(dt2$plot)!="factor") dt2$plot<- as.factor(dt2$plot)
# if (class(dt2$sec)!="factor") dt2$sec<- as.factor(dt2$sec)
# #f (class(dt2$tree)!="factor") dt2$tree<- as.factor(dt2$tree) # don't really want tree number to be a factor
# if (class(dt2$species)!="factor") dt2$species<- as.factor(dt2$species)
# if (class(dt2$dbht0)=="factor") dt2$dbht0 <-as.numeric(levels(dt2$dbht0))[as.integer(dt2$dbht0) ]               
# if (class(dt2$dbht0)=="character") dt2$dbht0 <-as.numeric(dt2$dbht0)
# if (class(dt2$cc)!="factor") dt2$cc<- as.factor(dt2$cc)
# if (class(dt2$condt0)!="factor") dt2$condt0<- as.factor(dt2$condt0)
# if (class(dt2$xcoord)=="factor") dt2$xcoord <-as.numeric(levels(dt2$xcoord))[as.integer(dt2$xcoord) ]               
# if (class(dt2$xcoord)=="character") dt2$xcoord <-as.numeric(dt2$xcoord)
# if (class(dt2$ycoord)=="factor") dt2$ycoord <-as.numeric(levels(dt2$ycoord))[as.integer(dt2$ycoord) ]               
# if (class(dt2$ycoord)=="character") dt2$ycoord <-as.numeric(dt2$ycoord)
# if (class(dt2$zelev)=="factor") dt2$zelev <-as.numeric(levels(dt2$zelev))[as.integer(dt2$zelev) ]               
# if (class(dt2$zelev)=="character") dt2$zelev <-as.numeric(dt2$zelev)
# if (class(dt2$elev0)!="factor") dt2$elev0<- as.factor(dt2$elev0)
# if (class(dt2$markmanip)!="factor") dt2$markmanip<- as.factor(dt2$markmanip)
# if (class(dt2$actualmanip)!="factor") dt2$actualmanip<- as.factor(dt2$actualmanip)
# if (class(dt2$cond06)!="factor") dt2$cond06<- as.factor(dt2$cond06)
# if (class(dt2$dam06)!="factor") dt2$dam06<- as.factor(dt2$dam06)
# if (class(dt2$sprt06)!="factor") dt2$sprt06<- as.factor(dt2$sprt06)
# if (class(dt2$notes1)!="factor") dt2$notes1<- as.factor(dt2$notes1)
# if (class(dt2$notes06)!="factor") dt2$notes06<- as.factor(dt2$notes06)
# if (class(dt2$dbh09)=="factor") dt2$dbh09 <-as.numeric(levels(dt2$dbh09))[as.integer(dt2$dbh09) ]               
# if (class(dt2$dbh09)=="character") dt2$dbh09 <-as.numeric(dt2$dbh09)
# if (class(dt2$cond09)!="factor") dt2$cond09<- as.factor(dt2$cond09)
# if (class(dt2$dam09)!="factor") dt2$dam09<- as.factor(dt2$dam09)
# if (class(dt2$sprt09)!="factor") dt2$sprt09<- as.factor(dt2$sprt09)
# if (class(dt2$notes09)!="factor") dt2$notes09<- as.factor(dt2$notes09)
# if (class(dt2$hwab109)!="factor") dt2$hwab109<- as.factor(dt2$hwab109)
# if (class(dt2$hwab209)!="factor") dt2$hwab209<- as.factor(dt2$hwab209)
# if (class(dt2$dbh14)=="factor") dt2$dbh14 <-as.numeric(levels(dt2$dbh14))[as.integer(dt2$dbh14) ]               
# if (class(dt2$dbh14)=="character") dt2$dbh14 <-as.numeric(dt2$dbh14)
# if (class(dt2$cond14)!="factor") dt2$cond14<- as.factor(dt2$cond14)
# if (class(dt2$dam14)!="factor") dt2$dam14<- as.factor(dt2$dam14)
# if (class(dt2$sprt14)!="factor") dt2$sprt14<- as.factor(dt2$sprt14)
# if (class(dt2$vigor14)!="factor") dt2$vigor14<- as.factor(dt2$vigor14)
# if (class(dt2$hwab114)!="factor") dt2$hwab114<- as.factor(dt2$hwab114)
# if (class(dt2$hwab214)!="factor") dt2$hwab214<- as.factor(dt2$hwab214)
# if (class(dt2$scale14)!="factor") dt2$scale14<- as.factor(dt2$scale14)
# if (class(dt2$note14)!="factor") dt2$note14<- as.factor(dt2$note14)
# if (class(dt2$dbh19)=="factor") dt2$dbh19 <-as.numeric(levels(dt2$dbh19))[as.integer(dt2$dbh19) ]               
# if (class(dt2$dbh19)=="character") dt2$dbh19 <-as.numeric(dt2$dbh19)
# if (class(dt2$cond19)!="factor") dt2$cond19<- as.factor(dt2$cond19)
# if (class(dt2$dam19)!="factor") dt2$dam19<- as.factor(dt2$dam19)
# if (class(dt2$sprt19)!="factor") dt2$sprt19<- as.factor(dt2$sprt19)
# if (class(dt2$vigor19)!="factor") dt2$vigor19<- as.factor(dt2$vigor19)
# if (class(dt2$hwab119)!="factor") dt2$hwab119<- as.factor(dt2$hwab119)
# if (class(dt2$hwab219)!="factor") dt2$hwab219<- as.factor(dt2$hwab219)
# if (class(dt2$scale19)!="factor") dt2$scale19<- as.factor(dt2$scale19)
# if (class(dt2$note19)!="factor") dt2$note19<- as.factor(dt2$note19)
# if (class(dt2$dbh24)=="factor") dt2$dbh24 <-as.numeric(levels(dt2$dbh24))[as.integer(dt2$dbh24) ]               
# if (class(dt2$dbh24)=="character") dt2$dbh24 <-as.numeric(dt2$dbh24)
# if (class(dt2$cond24)!="factor") dt2$cond24<- as.factor(dt2$cond24)
# if (class(dt2$dam24)!="factor") dt2$dam24<- as.factor(dt2$dam24)
# if (class(dt2$sprt24)!="factor") dt2$sprt24<- as.factor(dt2$sprt24)
# if (class(dt2$vigor24)!="factor") dt2$vigor24<- as.factor(dt2$vigor24)
# if (class(dt2$hwab124)!="factor") dt2$hwab124<- as.factor(dt2$hwab124)
# if (class(dt2$hwab224)!="factor") dt2$hwab224<- as.factor(dt2$hwab224)
# if (class(dt2$scale24)!="factor") dt2$scale24<- as.factor(dt2$scale24)
# if (class(dt2$note24)!="factor") dt2$note24<- as.factor(dt2$note24)
                
# Convert Missing Values to NA for non-dates
                
dt2$dbht0 <- ifelse((trimws(as.character(dt2$dbht0))==trimws("NA")),NA,dt2$dbht0)               
suppressWarnings(dt2$dbht0 <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt2$dbht0))==as.character(as.numeric("NA"))),NA,dt2$dbht0))
dt2$xcoord <- ifelse((trimws(as.character(dt2$xcoord))==trimws("NA")),NA,dt2$xcoord)               
suppressWarnings(dt2$xcoord <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt2$xcoord))==as.character(as.numeric("NA"))),NA,dt2$xcoord))
dt2$ycoord <- ifelse((trimws(as.character(dt2$ycoord))==trimws("NA")),NA,dt2$ycoord)               
suppressWarnings(dt2$ycoord <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt2$ycoord))==as.character(as.numeric("NA"))),NA,dt2$ycoord))
dt2$zelev <- ifelse((trimws(as.character(dt2$zelev))==trimws("NA")),NA,dt2$zelev)               
suppressWarnings(dt2$zelev <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt2$zelev))==as.character(as.numeric("NA"))),NA,dt2$zelev))
dt2$elev0 <- as.factor(ifelse((trimws(as.character(dt2$elev0))==trimws("NA")),NA,as.character(dt2$elev0)))
dt2$dbh09 <- ifelse((trimws(as.character(dt2$dbh09))==trimws("NA")),NA,dt2$dbh09)               
suppressWarnings(dt2$dbh09 <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt2$dbh09))==as.character(as.numeric("NA"))),NA,dt2$dbh09))
dt2$dbh14 <- ifelse((trimws(as.character(dt2$dbh14))==trimws("NA")),NA,dt2$dbh14)               
suppressWarnings(dt2$dbh14 <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt2$dbh14))==as.character(as.numeric("NA"))),NA,dt2$dbh14))
dt2$dbh19 <- ifelse((trimws(as.character(dt2$dbh19))==trimws("NA")),NA,dt2$dbh19)               
suppressWarnings(dt2$dbh19 <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt2$dbh19))==as.character(as.numeric("NA"))),NA,dt2$dbh19))
dt2$dbh24 <- ifelse((trimws(as.character(dt2$dbh24))==trimws("NA")),NA,dt2$dbh24)               
suppressWarnings(dt2$dbh24 <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt2$dbh24))==as.character(as.numeric("NA"))),NA,dt2$dbh24))
        
```

Clean up the tree data
```{r}
simes24<-dt2
simes24$species<-ifelse(simes24$species == "ACSA ", "ACSA", simes24$species)
simes24<-simes24 %>% mutate_at(c(2,3,4,7,15, 16, 17, 21, 22, 23, 28, 29, 30, 37, 38, 39, 46, 47, 48), as.factor)
summary(simes24)

#clean up improbable dbh values from 2004.
# 2004 harder to proof bc no prior measurements. Also, data were recorded into a voice recorder, which may have leant itself to some errors.
#trees 3606, 3902, 1901, 3920, and 3900 dbh in 2004 was more than 20cm greater than 2009 and subsequent meas. Change these to dbh09.
simes24$dbht0<-simes24$dbht0<-ifelse(simes24$tree %in% c(3902, 1901, 3920, 3900, 3606, 3911), simes24$dbh09, simes24$dbht0)
#trees 2721, 6701, 4173, 5300, and 3239 grew more than 10cm between 2004 and 2009. It's pretty clear what the correct value should be for 2004 for these.
simes24$dbht0<-simes24$dbht0<-ifelse(simes24$tree==2721, 57.0, simes24$dbht0) #dbht0 was recorded as 37.0
simes24$dbht0<-simes24$dbht0<-ifelse(simes24$tree==6701, 69.1, simes24$dbht0) #dbht0 was recorded as 59.1
simes24$dbht0<-simes24$dbht0<-ifelse(simes24$tree==4173, 67.0, simes24$dbht0) #dbht0 was recorded as 57.0
simes24$dbht0<-simes24$dbht0<-ifelse(simes24$tree==5300, 43.2, simes24$dbht0) #dbht0 was recorded as 33.2
simes24$dbht0<-simes24$dbht0<-ifelse(simes24$tree==3239, 48.5, simes24$dbht0) #dbht0 was recorded as 38.5

# exclude the 170 'off-HF' trees in plot 7 for entire study to look at change over time in trees consistently.
simes7off<-subset(simes24, simes24$note14 == 'off HF')
simes24<-subset(simes24, simes24$note14 != 'off HF'| is.na(simes24$note14)) #second part of this retains the NAs

##### calculate BA per tree, for live trees only - but why not wait until long format? #####
simes24$BAm204<-ifelse(simes24$condt0 %in% c('L', 'R'), (pi*((simes24$dbht0/200)^2)), NA)
simes24$BAm209<-ifelse(simes24$cond09 %in% c('L', 'R'), (pi*((simes24$dbh09/2)^200)), NA)
simes24$BAm214<-ifelse(simes24$cond14 %in% c('L', 'R'), (pi*((simes24$dbh14/2)^200)), NA)
simes24$BAm219<-ifelse(simes24$cond19 %in% c('L', 'R'), (pi*((simes24$dbh19/2)^200)), NA)
simes24$BAm224<-ifelse(simes24$cond24 %in% c('L', 'R'), (pi*((simes24$dbh24/2)^200)), NA)

#add plot size, block, and treatment
simes24$size.ha<-ifelse(simes24$plot==7, 0.6000, ifelse(simes24$plot %in% c(1,3),(90*85)/10000,ifelse(simes24$plot==2,(85^2)/10000,(90^2)/10000)))
simes24$block<-ifelse(simes24$plot %in% c('1', '2', '3', '8'),'valley', 'ridge')
simes24$trt<-ifelse(simes24$plot %in%c('1', '5'), 'girdled', ifelse(simes24$plot %in% c('2', '4'), 'logged', ifelse (simes24$plot %in% c('3', '6'), 'hemlock', 'hardwood')))
simes24$block<-as.factor(simes24$block)
simes24$trt<-as.factor(simes24$trt)
simes24$trt <- ordered(simes24$trt, levels = c("hemlock", "girdled", "logged", "hardwood"))

#SPECIES GROUPS
simes24 %>% count(species, sort=TRUE)
simes24$spgr<-ifelse(simes24$species == 'ACRU', 'red maple', ifelse(simes24$species  %in% c('BELE', 'BEAL', 'BEPA', 'BEPO'), 'birch', 
                                                                          ifelse(simes24$species=='PIST', 'white pine', 
                                                                          ifelse(simes24$species %in% c('QURU', 'QUVE','QUAL', 'QUBI'), 'oak',
                                        ifelse(simes24$species == 'TSCA', 'hemlock', 'other hardwood')))))

# add interval; cannot be zero :)
simes24$intervalt04<-NA
simes24$intervalt09<-ifelse(simes24$cond09 %in% c('L','D'), 2009-simes24$yeart0, NA)
simes24$intervalt09<-ifelse(simes24$intervalt09==0, 5, simes24$intervalt09)
simes24$intervalt141924<-5

# look at the live trees in 2024, can i code survivor vs recruit?
simes24$SR<-ifelse(simes24$tree>8000, 'Recruit', 'Survivor')
names(simes24)
simes24<-droplevels(simes24)
names(simes24)
levels(simes24$species)
simple24<-simes24 [c('plot', 'sec', 'xcoord', 'ycoord','trt', 'block', 'tree', 'species','spgr','size.ha', "yeart0", "SR",
                           'dbht0', 'dbh09', 'dbh14', 'dbh19', 'dbh24', 
                           'condt0', 'cond09', 'cond14', 'cond19', 'cond24')]
names(simple24)
tree_long<-reshape(simple24, direction="long", 
                   varying=list(dbh=c(13:17), cond=c(18:22)),
                   v.names = c("dbh", "cond"))

tree_long$year<-ifelse(tree_long$time==1, 2004, ifelse(tree_long$time==2, 2009, (ifelse(tree_long$time==3, 2014, ifelse(tree_long$time==4, 2019, 2024)))))
tree_long<-tree_long [c('plot', 'sec','xcoord', 'ycoord','year', 'trt', 'block', 'tree', 'species','spgr','SR','size.ha', 'yeart0' ,'dbh', 'cond' )]
summary(tree_long)
# take out cond=NA; these trees were dead in prior interval or not-yet recruited
tree_longNA<-subset(tree_long, is.na(tree_long$cond))
tree_long<-subset(tree_long, !is.na(tree_long$cond))
```

Step 1b. Pull out the recruits from just the center section of each plot to match the sample area used for the saplings, & exclude dead recruits
```{r}
recruitsLc<-subset(tree_long, tree_long$tree>8000 & tree_long$cond !='D' & tree_long$sec=='C')
recruitsLc$n.ha<-1/0.09 #divide by the size of the center section of the plot
RecruitDens<-recruitsLc %>% 
  group_by (plot, year, spgr) %>%
  summarise (density.ha=sum(n.ha))
```

Step 2a. Load the sapling data and organize it. 

Download sapling data from EDI, using the EDI's code generation in R tab. This code is copied from there
I subset this to read in only 106-05, the sapling data file (there are 7 datasets total, but the others aren't needed here)
```{r}
# Package ID: knb-lter-hfr.106.35 Cataloging System:https://pasta.edirepository.org.
# Data set title: Understory Vegetation in Hemlock Removal Experiment at Harvard Forest since 2003.
# Data set creator:  Aaron Ellison -  
# Data set creator:  Audrey Barker Plotkin -  
# Contact:    - Information Manager Harvard Forest  - hf-im@lists.fas.harvard.edu
# Stylesheet v2.16 for metadata conversion into program: John H. Porter, Univ. Virginia, jporter@virginia.edu      
# Uncomment the following lines to have R clear previous work, or set a working directory
# rm(list=ls())      

# setwd("C:/users/my_name/my_dir")       


inUrl5  <- "https://pasta.lternet.edu/package/data/eml/knb-lter-hfr/106/35/8309eef9e280907d910123c66e09d142" 
infile5 <- tempfile()
try(download.file(inUrl5,infile5,method="curl",extra=paste0(' -A "',getOption("HTTPUserAgent"),'"')))
if (is.na(file.size(infile5))) download.file(inUrl5,infile5,method="auto")

                   
 dt5 <-read.csv(infile5,header=F 
          ,skip=1
            ,sep=","  
        , col.names=c(
                    "plot",     
                    "year",     
                    "species",     
                    "in_fence",     
                    "out_fence",     
                    "no_info",     
                    "total",     
                    "note"    ), check.names=TRUE)
               
unlink(infile5)
		    
# Fix any interval or ratio columns mistakenly read in as nominal and nominal columns read as numeric or dates read as strings
# ABP: DON'T WANT SPECIES A FACTOR YET
                
if (class(dt5$plot)!="factor") dt5$plot<- as.factor(dt5$plot)
#if (class(dt5$species)!="factor") dt5$species<- as.factor(dt5$species)
if (class(dt5$in_fence)=="factor") dt5$in_fence <-as.numeric(levels(dt5$in_fence))[as.integer(dt5$in_fence) ]               
if (class(dt5$in_fence)=="character") dt5$in_fence <-as.numeric(dt5$in_fence)
if (class(dt5$out_fence)=="factor") dt5$out_fence <-as.numeric(levels(dt5$out_fence))[as.integer(dt5$out_fence) ]               
if (class(dt5$out_fence)=="character") dt5$out_fence <-as.numeric(dt5$out_fence)
if (class(dt5$no_info)=="factor") dt5$no_info <-as.numeric(levels(dt5$no_info))[as.integer(dt5$no_info) ]               
if (class(dt5$no_info)=="character") dt5$no_info <-as.numeric(dt5$no_info)
if (class(dt5$total)=="factor") dt5$total <-as.numeric(levels(dt5$total))[as.integer(dt5$total) ]               
if (class(dt5$total)=="character") dt5$total <-as.numeric(dt5$total)
if (class(dt5$note)!="factor") dt5$note<- as.factor(dt5$note)
                
# Convert Missing Values to NA for non-dates
                
dt5$in_fence <- ifelse((trimws(as.character(dt5$in_fence))==trimws("NA")),NA,dt5$in_fence)               
suppressWarnings(dt5$in_fence <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt5$in_fence))==as.character(as.numeric("NA"))),NA,dt5$in_fence))
dt5$out_fence <- ifelse((trimws(as.character(dt5$out_fence))==trimws("NA")),NA,dt5$out_fence)               
suppressWarnings(dt5$out_fence <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt5$out_fence))==as.character(as.numeric("NA"))),NA,dt5$out_fence))
dt5$no_info <- ifelse((trimws(as.character(dt5$no_info))==trimws("NA")),NA,dt5$no_info)               
suppressWarnings(dt5$no_info <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt5$no_info))==as.character(as.numeric("NA"))),NA,dt5$no_info))
dt5$total <- ifelse((trimws(as.character(dt5$total))==trimws("NA")),NA,dt5$total)               
suppressWarnings(dt5$total <- ifelse(!is.na(as.numeric("NA")) & (trimws(as.character(dt5$total))==as.character(as.numeric("NA"))),NA,dt5$total))
```

clean up the sapling data and subset to include: logged & girdled treatments only; hemlock (foundation) & birch (competitor) only
```{r}
saplings<-dt5
str(saplings)
saplings$species<-ifelse(is.na(saplings$species), 'none', saplings$species)

#### Data Wrangling ####
# group trees/saplings as hemlock (initial dominant), birch (post-disturbance dominant), and other
saplings$spgr<-ifelse(saplings$species == 'ACRU', 'red maple', ifelse(saplings$species  %in% c('BELE', 'BEAL', 'BEPA', 'BEPO'), 'birch', 
                                                                          ifelse(saplings$species=='PIST', 'white pine', 
                                                                          ifelse(saplings$species %in% c('QURU', 'QUVE','QUAL', 'QUBI'), 'oak',
                                        ifelse(saplings$species == 'TSCA', 'hemlock', 'other hardwood')))))


saplings$n.ha<-saplings$total/0.09 #divide by the size of the center section of the plot
SapDens<-saplings %>% 
  group_by (plot, year, spgr) %>%
  summarise (density.ha=sum(n.ha))

SapDens1<-subset(SapDens, SapDens$plot %in% c(1,2,4,5) & SapDens$spgr %in% c('birch', 'hemlock'))

```

Need to make sure there are zero values for each year/plot/spgr combo (80 rows = 10 year * 2 spgr * 4 plots)
```{r}

plot<-rep(c(1,2,4,5), each=20)
year<-rep(c(2004, 2007, 2009, 2011, 2013, 2015, 2017, 2019, 2021, 2023), each=2)
spgr<-rep(c('birch', 'hemlock'), times = 20)
saprow<-data.frame(plot, year, spgr)

```


Combine saplings and recruits to get a density for each of the years (recruits will be bumpy. . . hmm. should I just use '09, '14, '19 and '24?)
maybe also just use saplings as the metric so have more years? Let's try both

First, just saplings. The outcome (SapDens3) is ready to analyze!
```{r}

SapDens3<-merge(saprow, SapDens1, by = c('plot', 'spgr', 'year'), all.x = TRUE)
SapDens3[is.na(SapDens3)] <- 0
SapDens3$block<-ifelse(SapDens3$plot %in% c('1', '2', '3', '8'),'valley', 'ridge')
SapDens3$trt<-ifelse(SapDens3$plot %in%c('1', '5'), 'girdled', ifelse(SapDens3$plot %in% c('2', '4'), 'logged', ifelse (SapDens3$plot %in% c('3', '6'), 'hemlock', 'hardwood')))

```

Here, combine recruits and saplings, and then subset so just the girdled and logged plots, and just birch and hemlock
The recruit and sapling measurement years differed a bit for some of the years. Recruits: 2009, 2014, 2019, 2024; Saplings: 2004, 2007, 2009, 2011, 2013, 2015, 2017, 2019, 2021, 2023. Change the recruit year 2014 to 2013 and recruit year 2024 to 2023 so that these line up. I don't think this biases our data.
```{r}
#rename the year so I can later make a year column that matches up with the tree recruit sample years
RecruitDens$yearR<-RecruitDens$year
#create another year column 
RecruitDens$year<-ifelse(RecruitDens$yearR == 2014, 2013, ifelse(RecruitDens$yearR == 2024, 2023, RecruitDens$yearR))
RecruitDens<-RecruitDens[,c(1:4)] # columns match order and names with RecruitDens

#combine
RecSap<-rbind(SapDens, RecruitDens)
#add up the number per ha of saplings & recruits by plot and year and spgr
RecSap1<-RecSap %>% 
  group_by (plot, year, spgr) %>%
  summarise (density.ha=sum(density.ha))

RecSap2<-subset(RecSap1, RecSap1$plot %in% c(1,2,4,5) & RecSap1$spgr %in% c('birch', 'hemlock'))

#get all the rows including zeroes
RecSap3<-merge(saprow, RecSap2, by = c('plot', 'spgr', 'year'), all.x = TRUE)
RecSap3[is.na(RecSap3)] <- 0
RecSap3$block<-ifelse(RecSap3$plot %in% c('1', '2', '3', '8'),'valley', 'ridge')
RecSap3$trt<-ifelse(RecSap3$plot %in%c('1', '5'), 'girdled', ifelse(RecSap3$plot %in% c('2', '4'), 'logged', ifelse (RecSap3$plot %in% c('3', '6'), 'hemlock', 'hardwood')))

```

look at sapling density by plot (in girdled and logged) over time
```{r}
ggplot(RecSap3, aes(x = year, y = density.ha, color = trt)) +
  geom_point() +
  geom_smooth() +
  labs(x = "Year",
       y = "Sapling + recruit density (no./ha)") +
  facet_wrap(~spgr) +
  theme_minimal()
```

Calculate a rate (stems/ha/year) by using the number of years after 2005 (the year the treatments occurred and when stand initiation began) as the number to divide by to get stems/ha/year. And exclude 2004, because that is pre-treatment.
I considered the possibility that pre-existing saplings (those present in 2004) could muddy the results since they were there prior to any influence of the material legacies. However, that possibility is minimal: in most cases, any pre-existing saplings died between 2004 and 2007 so any new ones were after the treatment OR there were no saplings pre-treatment so any new ones were after the treatement. The exceptions to this were for hemlock saplings in the 2 logged plots: in these, there were existing hemlock saplings in 2004 and fewer (but not-zero) in 2007. If this introduces any error, it would be in exaggerating the response of the foundation species to the removal of the material legacy, which would make our conclusion that material legacy favors conspecific sapling recruitment more conservative.
```{r}
RecSap3$density.ha.yr<-RecSap3$density.ha/(RecSap3$year-2005)
RecSap3<-subset(RecSap3, RecSap3$year>2004)
```

save the data for the analysis part 'HemRemKaiAudreyFeb26.R'
```{r}
write.csv(RecSap3, 'HF-hemlock-birch-saps-per-year.csv', row.names=FALSE)
```


NOT DONE HERE.
Bring in the dead wood data. Revise this if needed. For now, we are just using the girdled vs logged treatment as a with/without material legacy
```{r}

# Dead wood sampled in 2005, 2007, 2009, 2011, 2013, 2015, 2017, 2021 (next 2025).
# Dead wood amounts inferred for 2004, 2014, and 2019 (to line up with tree census dates)
predictorsAll<-read.csv('DeadByPlot.csv', header=TRUE)
str(predictorsAll)
predictorsAll<-predictorsAll %>% mutate_at(c(2,3,5), as.factor)

# Trees (>5 cm dbh) measured in 2004, 2009, 2014, 2019, 2024
# Sapling trees (>1.3 m tall but <5 cm dbh) measured 2004, 2007, 2009, 2011, 2013, 2015, 2017, 2019, 2021, 2023
```

